A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-02-01 analysis/paper/paper.Rmd
Set up an initial pass at Brunlanes

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
library(stringr)
quickSetupOxcal()
```

@@

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English). Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @berg-hansen2009; @bjerck1990; @mjærum2018; @solheim2018]. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 kyr BCE (data from Hughes et al. 2016). The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center"}

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.

@@

# Data

To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b, @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

@@

```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset et al. 2018), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13}
bboxsites <- st_bbox(sites_sa)
bboxsites[1] <- bboxsites[1] - 15000
bboxsites[3] <- bboxsites[3] + 15000
bboxsites[2] <- bboxsites[2] - 5000
bboxsites[4] <- bboxsites[4] + 5000
bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017). A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```

@@

# Create plot showing the year drawn from the posterior sum
datsamp <- ggplot(posteriorprobs) +
  ggridges::geom_ridgeline(aes(x = dates, y = name,
                               height = probabilities * 50),
                           fill = "lightgrey") +
  # Create segment to indicate which year was drawn
  geom_segment(aes(x = sdate, xend = sdate, y = as.numeric(as.factor(name)),
                   yend = as.numeric(as.factor(name)) +
                     (posteriorprobs[round(dates) == sdate,
                                     "probabilities"]) * 50),
               colour = "red") +
  labs(x = "cal BCE")  +
  #scale_y_discrete(expand = c(0.01, 0)) +
  theme_bw() + theme(
    legend.position = "none",
    axis.line = element_line("black"),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_x_continuous(breaks = sort(c(-8200, -8000, sdate, -7600)))

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year from the posterior density esimate.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1. The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation. The second column displays the result of simulating the raised sea-level 1000 times. The more opaque the colour, the more times the sea-level was simulated to that location. The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px"}
knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
```

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m. However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to m below the site location. This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance. This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites.

An exponential function has been fit to the distribution. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. The sites treated here have been given a shoreline date using these values in Figure Figure \@ref(fig:shoredate).

@@

```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured. Results associated with agricultural acitivites are in grey. The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results.png"))

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory. Negative values have been set to zero. The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating  sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. They have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @glorstad2016; @bang-andersen2012]. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and neglible adjustments were therefore needed. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations. This information is not integrated in the approach outlined here, but could justify reducing the upper bound of the uncertainty ranges, especially for Pauler 1 and Pauler 2. While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration does speak further to the general applicability of the approach to shoreline dating outlined in this study. This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. This makes less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).

>>>>

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
library(stringr)
quickSetupOxcal()

# Set up so that I can call floatbarrier using regfloat=TRUE in the chunk 
# headings. See: 
# https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758
knitr::knit_hooks$set(plot = function (x, options) {
  float_correct <- function(f, y, opts)  {
    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
    else
      f(y, opts)
  }
  if (!is.null(options$out.width) || !is.null(options$out.height) ||
      !is.null(options$out.extra) || options$fig.align != "default" ||
      !is.null(options$fig.subcap)) {
    if (is.null(options$fig.scap))
      options$fig.scap = NA
    return(float_correct(knitr:::hook_plot_tex, x, options))
  }
  return(float_correct(knitr:::hook_plot_md_base, x, options))
})
```

@@

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English). Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @berg-hansen2009; @bjerck1990; @breivik2014; @solheim2018]. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 kyr BCE (data from Hughes et al. 2016). The study area defined later in the text is marked with a red outline.", out.width = "400px", fig.align="center", regfloat=TRUE}

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few, and are typically conducted on a site-by-site or project-by-project basis, meaning the basis for assuming shore-bound site location is largely a mosaic of smaller scale investigations.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b, @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

@@

```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset et al. 2018), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13, regfloat=TRUE}
bboxsites <- st_bbox(sites_sa)
bboxsites[1] <- bboxsites[1] - 15000
bboxsites[3] <- bboxsites[3] + 15000
bboxsites[2] <- bboxsites[2] - 5000
bboxsites[4] <- bboxsites[4] + 5000
bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017). A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```

@@

# Create plot showing the year drawn from the posterior sum
datsamp <- ggplot(posteriorprobs) +
  ggridges::geom_ridgeline(aes(x = dates, y = name,
                               height = probabilities * 50),
                           fill = "lightgrey") +
  # Create segment to indicate which year was drawn
  geom_segment(aes(x = sdate, xend = sdate, y = as.numeric(as.factor(name)),
                   yend = as.numeric(as.factor(name)) +
                     (posteriorprobs[round(dates) == sdate,
                                     "probabilities"]) * 50),
               colour = "red") +
  labs(x = "cal BCE")  +
  #scale_y_discrete(expand = c(0.01, 0)) +
  theme_bw() + theme(
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_x_continuous(breaks = sort(c(-8200, -8000, sdate, -7600)))

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year from the posterior density esimate.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance mesaures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

```{r example-3, echo = FALSE, fig.cap = "The result of 1000 simulation runs for each of the two groups of dates on the site Hegna vest 1. The first column of plots shows the radiocarbon probability density estimate from where dates were drawn during simulation. The second column displays the result of simulating the raised sea-level 1000 times. The more opaque the colour, the more times the sea-level was simulated to that location. The third column shows violin plots of the different distance measures across all simulations.", out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_result.png"))
```

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m. However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. An exponential function has been fit to the distributions for vertical distance. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. Using the resulting exponential decay function, the sites treated here have been given a shoreline date in Figure \@ref(fig:shoredate).

@@

```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured. Results associated with agricultural acitivites are in grey. The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results.png"))

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory. Negative values have been set to zero. The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating  sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95\\% probability range for the shoreline date.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites are considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations. This information is not integrated in the approach outlined here, but could justify reducing the upper bound of the uncertainty ranges, especially for Pauler 1 and Pauler 2. While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration does speak further to the general applicability of the approach to shoreline dating outlined in this study. This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating the Brunlanes sites. A) The result of applying the approach to shoreline dating outlined here, plotted against the dates provided by Jaksland (2014) given in black. B) Map of the Pauler sites and Sky 1 with simulated sea-level using the probability density associated with the shoreline date for Pauler 1.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```
